On the reality of the quantum state once again: A no-go theorem for $\psi$-ontic models

### 2023-02-10 Articles/2021-HarriganSpekkens/HarriganSpekkensTakeTwo.tex
HarriganSpekkens: another pass

In this paper we are not going to rebut the theorem itself. We instead aim to carefully re-examine one of its premises, i.e.\ HS classification between $\psi$-ontic and $\psi$-epistemic ontological models on top of which the PBR result is derived. We believe that a careful analysis of this assumption forces us to re-evaluate the significance of this classification as well as the meaning of the PBR conclusion. In other words, we re-examine the definitions and features of such models in order to understand whether they constitute a sound basis from which to draw conclusions on the nature of $\psi$, given the influence that the HS framework has in quantum foundations. As the reader will see, our analysis suggests that they are not.

While the PBR theorem is a no-go result for $\psi$-epistemic models, in this essay we provide a no-go theorem for $\psi$-ontic models: using information theoretic considerations we show that $\psi$-ontic models cannot reproduce all the predictions, results and implications of QM since the lack of overlap of epistemic states---a necessary condition for a model to be $\psi$-ontic---requires that all states must be orthogonal, which openly contradicts quantum theory. Therefore, we conclude, $\psi$-ontic models should not be employed in order to classify quantum theories.

Combining our result with the PBR argument, we conclude that the HS classification itself is fundamentally empty, i.e.\ both $\psi$-epistemic and $\psi$-ontic models do contradict quantum mechanics. Consequently, HS framework should be employed neither to classify interpretations of the quantum formalism, nor to draw conclusion about the nature of $\psi$.

@@

To reproduce quantum theory, an ontological model has to reproduce not simply the probability of transitions during measurements but also all the results of quantum statistical mechanics and quantum information theory. Therefore, among other things, it has to obtain the correct values for entropy of mixed states. \textcolor{blue}{The natural way is by using standard information theoretic tools over the space of epistemic states which, as we will see, rules out $\psi$-ontic models. The degree of overlap between distributions, in fact, is not a choice, but is fixed by the value of entropy.} %@Gabriele: expand these points, i.e. explain more clearly the inference that you have in mind.

Let us first review a crucial property of the information entropy, both in the classical setting (i.e.\ the Shannon/Gibbs entropy) and in the quantum setting (i.e.\ the von Neumann entropy).\footnote{We show the explicit calculations in the appendix.} %@ Gabriele: why is it important to introduce both classical and quantum property of information entropy? Questi sono tutti commenti anti-referee. Esplicita meglio a cosa servono queste cose.

\textbf{Classical information entropy of mixed non-overlapping distributions}. Let $\rho_1$ and $\rho_2$ be two classical distributions over a space $\Lambda$ with measure $\lambda$. The entropy for each distribution is given by the usual formula\footnote{The logarithm is assumed to be in base 2.}, for example
\begin{equation}\label{shannon_entropy}
	H(\rho_1) = - \int_\Lambda \rho_1(\lambda) \log \rho_1(\lambda) d\lambda.
\end{equation}
Suppose the two distributions are disjoint, and let $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ be a uniform mixture of the two distributions. The entropy of $\rho$ is given by
\begin{equation}\label{entropy_nonoverlap}
	H(\rho) = 1 + \frac{1}{2} H(\rho_1) + \frac{1}{2} H(\rho_2).
\end{equation}
Note how the non-overlapping assumption fixes the entropy of the mixed state.

\textbf{Quantum information entropy of quantum mixed states}. Now suppose $\psi$ and $\phi$ are two pure quantum states and let $p = | \langle \psi | \phi \rangle |^2$ be the probability of transition from one to the other. Consider the mixed state $\rho = \frac{1}{2} | \psi \rangle \langle \psi | + \frac{1}{2} | \phi \rangle \langle \phi |$. Its entropy is given by
\begin{equation}\label{entropy_mixed}
	H(\rho) = H\left(\frac{1+\sqrt{p}}{2}, \frac{1-\sqrt{p}}{2}\right).
\end{equation}

@@

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping distributions over a space $\Lambda$,  \textcolor{blue}{therefore eq. \ref{entropy_nonoverlap} applies}. %@Gabriele: mi rendo conto che  stupido ma  un commento anti-referee. Tu dici che l'eq. 7 che riguarda distribuzioni classiche si applica al nostro caso che riguarda stati quantistici. Esplicita a prova di referee perchŽ si applica l'eq. 7 e non 8
Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
		H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
		H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
		&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. That is, we must have that
\begin{equation}\label{orthogonal}
	\langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

@@

We should also note that one has problems for $\psi$-epistemic models. While PBR leaves the possibility of $\psi$-epistemic models that use correlations, the correlations is also fixed by the entropy. %Commenti anti-reviewer: what do you mean here exactly when you speak about correlations? Do you know the existence of one of these models?
It is a well-known result in both classical and quantum information theory that \textcolor{red}{the entropy of a joint distribution of the marginal if and only if the subsystems are independent} (\cite{Ash:2010}, \cite{Nielsen:2010}). %@Gabriele: a verb seems to be missing in the red sentence.
Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This, combined with PBR, tells us that no models can reproduce quantum statistical mechanics and quantum information theory.

@@

In principle, we do not exclude that someone may find a different construction to reproduce the correct entropies. There are, however, at least two objective difficulties that cannot be circumvented. First of all, information entropy is strictly concave, meaning that $H(p \rho_1 + (1-p) \rho_2) \geq p H(\rho_1) + (1-p) H(\rho_2)$ with the equality valid only if $\rho_1$ and $\rho_2$ are the same exact state. \textcolor{blue}{Therefore entropy fixed equality}. %spiega meglio
Because the mapping between epistemic states and density matrices is not injective,\footnote{Quantum mixture do not have a single decomposition in pure states. For example, the maximally mixed state of a qubit can be achieved with either an equal mixture of spin up and down or an equal mixture of spin left and right.} epistemic states must form equivalence classes and the strict concavity must be lost. That is, one can make mixtures of different ensemble without increasing entropy.

Secondly, as equation \ref{entropy_mixed} shows, the compute of the entropy requires the inner product. Therefore, whatever other structure one puts on top of epistemic states, essentially redefines the inner product. In other words, the idea that the inner product just represents transition probabilities, which, in our view, is the key idea the ontological model uses to explain ``what really happens'' does not work.

It is worth stressing, however, that it is not our task to solve these problems since we are neither the proponents of this model, nor supporters of it. Our only aim is to underline that the HS framework, as it currently stand, is highly problematic to draw any conclusion about the interpretation of quantum theory and the nature of $\psi$.

@@

In this paper we proved that $\psi$-ontic models are not compatible with quantum information theory, and therefore with quantum mechanics itself.

One may think that the issue could be circumvented by substituting the Shannon entropy formula. %with what? 
That is, the problem would not be with the non-overlapping assumption made by HS but rather with how the entropy is calculated. However, the link between probability theory, information theory and measure theory is so tight that it makes this objection untenable. If we take a set of states $U$ with a measure $\mu(U)$ that quantifies the states in the set, the probability of each state for a uniform distribution is given by $\rho = 1 / \mu(U)$ and the entropy by $\log \mu(U)$. The measure used to count states, the probabilities assigned to states and the entropy assigned to distributions are not independent mathematical structures: failure of one entails the failure of all. We will look deeper into this issue in a later work.

Finally, combining our result with the PBR theorem it follows that both $\psi$-ontic and $\psi$-epistemic models cannot reproduce QM, contradicting the results and the predictions of quantum theory. This fact shows in turn that HS categorization of ontological models is fundamentally empty. Thus, such a framework should be employed neither to classify quantum interpretations, nor to draw conclusions on the nature of the quantum state.

Referring to this, being the HS definitions inadequate to reproduce QM, it follows that any result derived from them rests on a technically ill-defined basis, as for instance the PBR theorem itself. If this is true, then, it is not demonstrated (i) that quantum states must necessarily be HS-$\psi$-ontic, and (ii) that HS-$\psi$-epistemic interpretations are contradicted by quantum theory. Concluding, we believe that the PBR theorem does not make claims on the reality of the quantum state, but it rather shows that HS definition of $\psi$-epistemic model contradicts QM. If interpreted in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories. Argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.

>>>>

While the discussion has been focused mainly on whether $\psi$-epistemic models are problematic or whether the PBR argument itself is problematic, we ask a different question: what if the HS classification itself is problematic? In particular, the framework assumes that once a full description of the transition probability is given for pure states, mixed states are constructed simply as a classical statistical mixture. But classical mixtures and quantum mixtures have different properties. For example, classical mixtures have a single decomposition in terms of pure states while quantum mixtures do not. Studying the behavior of states under mixing, then, opens a new venue of possible violations that are not as easily seen by looking at transition probability.

The core insight is that the degree of overlap between two probability distribution affects how the entropy of their mixture relates to the entropy of the components. Consequently, we find that the lack of overlap required by $\psi$-ontic models forces all pure states to be orthogonal. Since this openly contradicts quantum theory, we conclude that $\psi$-ontic models are not viable.

Combining our result with the PBR argument, we conclude that the HS classification itself is fundamentally flawed: both $\psi$-epistemic and $\psi$-ontic models do contradict quantum mechanics.\footnote{In principle, PBR leaves open the possibility of $\psi$-epistemic models that violate statistical independence. Given that statistical independence is linked to entropy additivity, that loophole can be closed as well.} Consequently, HS framework should be employed neither to classify interpretations of the quantum formalism, nor to draw conclusion about the nature of $\psi$.

@@

The exposition of the HS classification makes it seem one has complete freedom in choosing the overlap between the overlap between epistemic states. In this section we will show that this is not the case: this is constrained by entropy of mixtures. These constraints cannot be replicated by a $\psi$-ontic model with standard information theoretic techniques. Since quantum statistical mechanics and quantum information theory depend on the correct value of entropy on mixtures, $\psi$-ontic models will violate quantum theory.

To show the violation, we will first study how the entropy behaves when mixing two states, both for the Shannon/Gibbs entropy and the von Neumann entropy.\footnote{Full calculations with all steps are included in the appendix.} The first will be used to calculate entropy of statistical mixtures of epistemic state, since equation \ref{epistemic_mixing} tells us the mixture is done according to a standard (classical) measure of probability. The second will be used to calculate the entropy on the quantum case.

\textbf{Information entropy of mixed non-overlapping probability distributions}. Let $\rho_1$ and $\rho_2$ be two probability density function over a space $\Lambda$ with measure $\lambda$. That is, they are the Radon-Nikodym derivative of the respective probability measures and the measure $\lambda$. The entropy for each distribution is given by the usual formula\footnote{The logarithm is assumed to be in base 2.}, for example
\begin{equation}\label{shannon_entropy}
	H(\rho_1) = - \int_\Lambda \rho_1(\lambda) \log \rho_1(\lambda) d\lambda.
\end{equation}
Suppose the two distributions are disjoint, and let $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ be a uniform mixture of the two distributions. More precisely, $\rho$ is the Radon-Nikodym derivative of the average of the two probability measures with respect to $\lambda$. The entropy of $\rho$ is given by
\begin{equation}\label{entropy_nonoverlap}
	H(\rho) = 1 + \frac{1}{2} H(\rho_1) + \frac{1}{2} H(\rho_2).
\end{equation}
Note how the non-overlapping assumption fixes the entropy of the mixed state.

\textbf{Quantum information entropy of quantum mixed states}. Now suppose $\psi$ and $\phi$ are two pure quantum states and let $p = | \langle \psi | \phi \rangle |^2$ be the probability of transition from one to the other. Consider the mixed state $\rho = \frac{1}{2} | \psi \rangle \langle \psi | + \frac{1}{2} | \phi \rangle \langle \phi |$. Its entropy is given by
\begin{equation}\label{entropy_mixed}
	H(\rho) = H\left(\frac{1+\sqrt{p}}{2}, \frac{1-\sqrt{p}}{2}\right).
\end{equation}
where the right side of the equation is the Shannon entropy calculated on the given values. Therefore the entropy of an equal mixture of two pure states depends only on the probability of measuring one having prepared the other.

@@

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping probability distributions over a space $\Lambda$. That is, they are the two probability density functions defined to be the Radon-Nikodym derivatives between the respective probability measures over $\Lambda$ and the measure $\lambda$. Equation \ref{epistemic_mixing} tells us that the probability measure combine in the usual way, and therefore an equal mixture of the two states must obey \ref{entropy_nonoverlap}. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states in quantum mechanics is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
		H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
		H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
		&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. In fact, the only case the Shannon entropy for a two element distribution is equal to one is for the uniform distribution $H(1/2, 1/2) = 1$.  That is, we must have that
\begin{equation}\label{orthogonal}
	\langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

@@

If we combine this result with PBR, the only models that are not ruled our are $\psi$-epistemic models where the objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also been closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

@@

\section{Throwing out information theory}

Note that our conclusion relies on the use of standard information theoretic tools. In principle, we do not exclude that someone may find a different construction to reproduce the correct entropies. That is, one may give a different definition of entropy to use on the ontic space. We believe this to be highly unlikely as there are, at least, two objective difficulties that cannot be circumvented.

First of all, information entropy is strictly concave, meaning that $H(p \rho_1 + (1-p) \rho_2) \geq p H(\rho_1) + (1-p) H(\rho_2)$ with the equality valid only if $\rho_1$ and $\rho_2$ are the same exact state. That is, we can check whether two states $\rho_1$ and $\rho_2$ are the same state simply by creating a mixture $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ and see whether we end up with the same entropy or not. In other words: \emph{entropy fixes equality}. %spiega meglio
The issue here is that the mapping between epistemic states and density matrices is not injective. Quantum mixtures, in fact, do not have a single decomposition in pure states. For example, the maximally mixed state of a qubit can be achieved with either an equal mixture of spin up and down or an equal mixture of spin left and right. Therefore the same quantum mixture must correspond to an equivalence class of epistemic states, which means whatever definition we have on entropy for epistemic states must not satisfy strict concavity to be able to agree with the quantum case.\footnote{The degree of degeneracy one has to require is also extreme. For a qubit, in the simplest case, we would have a $\psi$-ontic state for each direction of spin, meaning the space of ontic states is the sphere $S^2$. The space of all possible epistemic states is the set of all possible distribution integrable to one. This is an infinite dimensional function space. However, the space of mixed state for a quibit is the interior of a sphere, which is a bounded three-dimensional manifold. The equivalence classes of epistic states, then, must be infinite dimensional.} That is, one can make mixtures of different ensemble without increasing entropy, which does not make physical sense.

Secondly, as equation \ref{entropy_mixed} shows, computing the entropy requires the inner product. In fact, if we had the entropy for all even mixtures of all pairs of pure states, we could reconstruct the inner product. Therefore, whatever other structure one puts on top of epistemic states, essentially redefines the inner product. Also, it has to do it in a way compatible with the other definitions of the model, therefore this structure will likely be ad-hoc for each model. In other words, the idea that the inner product just represents transition probabilities, which, in our view, is the key idea the ontological model uses to explain ``what really happens'' does not work.

One, naturally, will object that we haven't strictly proven that any of this is impossible or that there is no reasonable way to do it. Granted. However, it is not up to us to show that HS models cannot be made to work outside of the scope of standard information theory. And in any case, we believe we put forth enough evidence that the HS framework has serious flaws which, at the very least, must be understood by those that use it to draw any conclusion about the interpretation of quantum theory and the nature of $\psi$.

@@

By using standard information theoretic techniques, we have shown that $\psi$-ontic models are not compatible with quantum information theory, and therefore with quantum mechanics itself. Given that the PBR theorem already ruled out $\psi$-epistemic models, the HS categorization of ontological models is fundamentally empty. Thus, such a framework should be employed neither to classify quantum interpretations, nor to draw conclusions on the nature of the quantum state.

The key problem of the HS categorization is that it is ``not quantum enough''. The framework tries to hide the non-classicality at the level of each ontic state, which, by construction, cannot be decomposed further using a standard probability measure. However, it posits that statistical mixtures are simply classical mixture in the sense that they are captured by a standard probability measure. But quantum mixtures are non-classical in the same way that the quantum transition probabilities defined by the inner product are non-classical. Therefore the HS categorization prevents one to describe quantum statistical mixing.

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable not because there is a problem with $\psi$-epistemic models in particular, but rather there is an underlying problem in the HS definitions. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories. Argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.
